**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**

### 2019-08-02 submission/manuscript.Rmd
changes before meeting 2 with Pat

---
title: '**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**'
csl: mbio.csl
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  word_document: default
geometry: margin=1.0in
bibliography: references.bib
---

@@

__Predictive performance and generalizability of the seven models.__

We evaluated the predictive performances of seven binary classification models when applied to held-out test data using the AUROC metric [Figure 2]. Random forest had significantly higher test AUROC values than the other models for detecting SRNs when AUROC values were compared to the other six by Wilcoxon rank sum test (p < 0.01). The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`). L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another. They had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`) respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2]. Random forest had the highest median AUROC for detecting SRN. Despite its simplicity, the L2-regularized logistic regression was second best in predictive performance.

@@

__Interpretation of each ML model.__

Interpretability is the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017]. Because we often use ML models not just to predict a health outcome but also to learn the ecology behind a disease, model interpretation becomes crucial for microbiome studies. The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost decrease in interpretability as they increase in complexity.  In this study we highlighted two methods to interpret models with varying complexity.

>>>>

---
title: '**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**'
csl: mbio.csl
fontsize: 11pt
output:
  word_document: default
geometry: margin=1.0in
bibliography: references.bib
---

@@

__Predictive performance and generalizability of the seven models.__ We evaluated the predictive performances of seven binary classification models when applied to held-out test data using the AUROC metric [Figure 2]. Random forest had significantly higher test AUROC values than the other models for detecting SRNs when AUROC values were compared to the other six by Wilcoxon rank sum test (p < 0.01). The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`). L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another. They had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`) respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2]. Random forest had the highest median AUROC for detecting SRN. Despite its simplicity, the L2-regularized logistic regression was second best in predictive performance.

@@

__Interpretation of each ML model.__ Interpretability is the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017]. Because we often use ML models not just to predict a health outcome but also to learn the ecology behind a disease, model interpretation becomes crucial for microbiome studies. The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost decrease in interpretability as they increase in complexity.  In this study we highlighted two methods to interpret models with varying complexity.
